---
title: "AJB2021_FVT_CEPF_S2"
author: "Robert L. Baker"
email: "robert.baker@miamioh.edu"
affiliation: "https://rlbakerlab.com"
date: "12/29/2020"
abstract: "This document includes links to an example dataset and code for employing an integral approach to function valued trait data."
tags: [R, FVT, Brassica, AJB, Function-Valued Trait, American Journal of Botany, Development, High throughput phenotyping, HTP]
output: html_document
---
#### Background and explanation
The data used here are from Li et al, 2020. "Epistatic Transcription Factor Networks Differentially Modulate Arabidopsis Growth and Defense" Genetics vol. 214 no. 2 529-541; https://doi.org/10.1534/genetics.119.302996.

The authors have time-series data on plant growth. When they model these data they find that the growth of different genotypes is best fit by different functions. Therefore, the FVT approach taken where a function is fit to data and parameters of that function are used "as data" won't work because the functions are different."

Rather than using the parameters of an individual function "as data" we find the area under the curve described by each function and use that area "as data".  

We run a couple of quick correlations test to see whether area under the curve is correlated with early as well as late time points.  Not unsurprisingly, area under the curve is very strongly correlated with late developmental time points. However, diff is also significantly and strongly correlated with early developmental time points.

It should be noted that the data used here are LS means for each day rather than replicate level data. Depending on the goal of the analysis, it is equally possible to find the area under the curve for individual replicates and then calculate LS means for the area under the curve. 

```{r startup, results='hide'}
rm(list=ls()) #remove all objects from the R environment 
gc() #can help return memory to R after large objects have been called.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
r<-getOption("repos")
r["CRAN"]<-"https://cloud.r-project.org"
options(repos= r)

if(!require(tidyverse)){
  install.packages("tidyverse", dependencies=TRUE)
}
if(!require(readxl)){
  install.packages("readxl", dependencies=TRUE)
}
if(!require(filesstrings)){
  install.packages("filesstrings", dependencies=TRUE)
}
```
```{r, loadpackages, message=FALSE}
library(tidyverse)
library(readxl)
library(filesstrings)
```

#### Download the files, unzip & remove whitespace from filenames:
```{r, delete_whitespace, eval=FALSE}
#set working directory & load data:
#download zipped supplimental files at: https://gsajournals.figshare.com/ndownloader/files/20239863
#unzip files and set working directory to the folder "File S1" (or wherever you have saved the unzipped files)

remove_filename_spaces(replacement = "_") #replaces all whitespace in filenames with "_"
```

#### Load files and generate a new working dataframe.
The excel files contains 2 rectangles of data from separate growth chambers and these will need to be loaded separately.
```{r, load files, message=FALSE}
growth_CEF<-read_excel("Supplemental_Data_Set_1,_lsmeans_of_growth,_flowering_and_GLS_for_single_mutants.xlsx", range="B6:N27")
growth_CEF<-cbind("CEF", growth_CEF[,c(1,3:length(growth_CEF))])
colnames(growth_CEF)[1]<-"chamber"

growth_LSA<-read_excel("Supplemental_Data_Set_1,_lsmeans_of_growth,_flowering_and_GLS_for_single_mutants.xlsx", range="B32:N53")
growth_LSA<-cbind("LSA", growth_LSA[,c(1,3:length(growth_LSA))])
colnames(growth_LSA)[1]<-"chamber"

#generate a single dataframe with data from both growth chambers.
growth<-rbind(growth_CEF, growth_LSA)

#pull out growth data and store it as a list.
growth.list<-as.list(as.data.frame(t(growth[,4:13])))
```

#### Test case with graphical interpretation. 
Here, we use just the first line of data to calculate the area under the curve, which we designate "diff".
```{r testcase}
y1<-growth.list$V1 #grab growth data from the first line of the data set
y2<-rep(0, length(y1)) #set a baseline, in this case zero (but you could also compare two functions)
x<-c(9,11,13,15,17,19,21,23,25,27) #x-axis time values that correspond to y-axis growth data; hard coded for simplicity

f1<-approxfun(x,y1-y2) #approximate the function of the relationship between growth and time
f2<-function(z)abs(f1(z)) #absolute value of the difference between the baseline value and the growth data in question. The absolute value doesn't make much difference with baseline of zero but might be important if you wanted to directly compare 2 genotypes whose growth function intersected at some point - or a single genotype in multiple environments with different growth functions that intersected. 

plot(x,y1,type="l",ylim=c(0,40), ylab="growth", xlab="Time (days)", main="Growth Curve") #plot the estimated function

lines(x,y2,type="l",col="red") #plot the "baseline" value - zero in this case
polygon(c(x,rev(x)),c(y1,rev(y2)), col="red") #fit a polygon to the area between the function and the baseline: in this case the polygon represents the growth that the first individual underwent over time.

diff<-integrate(f2,min(x),max(x)) #do some calculus to find the area of that polygon

diff
```


#### Do this for the entire dataset (but skip the plots):
```{r loop}
diff<-NULL #set up a couple of empty variables
err<-NULL #set up a couple of empty variables
for(i in 1:nrow(growth)){
	y1<-get(paste("V", i, sep=""), growth.list)
	y2<-rep(0, length(y1))
	x<-c(9,11,13,15,17,19,21,23,25,27)
	
	f1<-approxfun(x,y1-y2)
	f2<-function(z)abs(f1(z))
	dif<-integrate(f2,min(x), max(x))
	
	diff[length(diff)+1]<-dif$value #append the current differential to "diff"
	err[length(err)+1]<-dif$abs.error	#append the current error estimate to "err"
	}

#add the two new variables, diff and err to the datset
dat1<-cbind(growth, diff, err)
```


#### Run a couple very simple correlation tests to compare the "area under the curve" (diff) to individual values used to generate the curve (first and last values):
```{r cortest}
cor.test(dat1$day27, dat1$diff) # test the correlation between the area under the function-value-trait curve and the last day of data collection.

#Diff is correlated to that final value. What about the first value? We would expect that contributes relatively little to the area under the curve, based on the figure above.

cor.test(dat1$day9, dat1$diff)

#The correlation is, as expected, not as strong. But an r of 0.762 is still pretty good, and it is highly significant.
```

#### How well does the "area under the curve" (integral) apporach correlate with the slopes calculated in Li et al 2020?
```{r cortest2}
cor.test(dat1$slope, dat1$diff)
#The correlation is not as strong as with individual datapoints, but is still substantial and certainly significant.
```

#### Citations and documentation
```{r, citations, echo=FALSE}
cite<-citation()[1]
print(cite, bibtex=FALSE)
shortRversion <- function() {
   rvs <- R.version.string
   if(grepl("devel", (st <- R.version$status)))
       rvs <- sub(paste0(" ",st," "), "-devel_", rvs, fixed=TRUE)
   gsub("[()]", "", gsub(" ", "_", sub(" version ", "-", rvs)))
}
shortRversion()
tidy<-citation("tidyverse")
print(tidy, bibtex=FALSE)
string<-citation("filesstrings")
print(string, bibtex=FALSE)
read<-citation("readxl")
print(read, bibtex=FALSE)
```